DRAFT -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- DRAFT

### 2024-03-22 docs/_tex/index.tex
pierre's comments

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
seagrass decline and habitat fragmentation. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the use of satellite data within monitoring programs. Yet,
satellite remote sensing struggles with spatial and spectral resolution,
making it difficult to discriminate seagrass from other macrophytes in
highly heterogenous meadows. Drone images at a very high spatial
resolution offer a promising solution to address challenges related to
spatial heterogeneity and intrapixel mixture. This study focuses on
using drone acquisitions with a ten spectral band sensor mirroring those
of Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12m and 120m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. This work showed that although they
share the same pigmentary composition, seagrass and green algae can be
discriminated using a multispectral sensor with ten spectral bands at
444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. The algorithm
should now be adapted for Sentinel-2 to see if this discrimination is
still operational at a coarser spatial resolution.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between the land and oceans \citep{unsworth2022}. Seagrass meadows
provide a myriad of ecosystem services to humanity, including carbon
sequestration, oxygen production, protection against sea-level rise and
coastline erosion, and limitation of eutrophication. They serve as vital
habitats for a diverse array of marine and terrestrial species,
providing living, breeding, and feeding grounds \citetext{\citealp[
]{gardner2018}; \citealp[ ]{Zoffoli2022}; \citealp{jankowska2019}}. Due
to the concentration of human activities in coastal zones, seagrass
meadows are directly exposed to and impacted by anthropogenic pressures.
Global regression and fragmentation are currently observed due to
climate change, diseases, urbanization, land reclamation, dredging,
competition with alien species, and reduction in water quality
\citetext{\citealp[ ]{nguyen2021}; \citealp[ ]{soissons2018}; \citealp[
]{orth2006}; \citealp[ ]{lin2018}; \citealp[ ]{duffy2019}; \citealp[
]{rasheed2011long}; \citealp{chefaoui2018dramatic}}. Both habitat
fragmentation and reduction, in turn, can severely copromise the
effectiveness of ecosystem services provided by seagrass meadows. While
improvements in water quality have been recently reported in Europe,
allowing an overall recovery of seagrass ecosystems at local scale, many
coastal waters worldwide are still subjected to strong eutrophication
processes \citetext{\citealp[ ]{deSantos2019}; \citealp{Zoffoli2021}}.
Coastal eutrophication has been ass ociated to excessive accumulation of
green macroalgae, so-called green tides (\citep{devlin2023nutrients}).
Green tides produce shade and suffocation over seagrass individuals,
thus threatening the health of seagrass ecosystems \citep{wang2022}.

@@

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a
coarse spatial resolution (\textgreater100m) over these restricted
extension areas. Missions with a high spatial resolution such as
Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral
resolution. The limited number of spectral bands is indeed a challenge
to accurately discriminate seagrass from others co-existing macrophytes.
In particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However the application of this
approach to satellite RS remains to be validated. Moreover patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ
measurements, due to their ability to provide spatially-explicit
observations at very high spatial resolutions (pixel size from cm to mm)
while capturing data at multi-spectral resolution \citetext{\citealp[
]{fairley2022drone}; \citealp{oh2017use}}. The versatility of drones
allows for their application across a diverse thematic range , from
coastal zone management \citetext{\citealp[ ]{adade2021}; \citealp[
]{casella2020}; \citealp{angnuureng2022}} to mapping species
distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[
]{Roman2021}; \citealp{Brunier2022Topographic}}. However, when applied
to coastal habitat mapping, most case studies are limited to a single
flight, restricting the generalizability of their application over wider
geographical scales \citetext{\citealp[ ]{Roman2021}; \citealp[
]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes, with a particular focus on discriminating
Magnoliopsida and Chlorophyceae (Seagrass and Green Algae respectively).
Ten drone flights were performed over soft-bottom intertidal areas along
two European countries (France and Portugal), covering a wide range of
habitats, from monospecific seagrass meadows to meadows mixed with
green, or red macroalgae. A deep learning algorithm was trained and
validated for macrophyte discrimination, emphasizing applicability
across diverse sites without a loss of prediction accuracy.

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourngeuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Green represents
intertidal areas.}

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take 1.2
million pixel reflectance photographs with ten spectral bands ranging
from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668,
705, 717, 740 and 840 nm. To ensure consistent lighting conditions
across flight paths, the drone's trajectory was aligned to maintain a
solar azimuth angle of 90 degrees. An overlap of 70\% and 80\% (side and
front respectively) between each image was set for each flight. A
downwelling light sensor (DLS2) was used to acquire irradiance data
concomitantly with the camera measurements. Raw data were calibrated in
reflectance using a calibration panel reflective at \textasciitilde50\%
provided by the manufacturer. Across all sites, flights were made at two
different altitudes : 12 m or/and 120 m. (Table~\ref{tbl-flights}).

@@

\caption{\label{tbl-flights}List of drone flight, summarising the date,
the altitude and the purpose of each flight. 12 m and 120 m flights have
a spatial resolution of 8 and 80 mm respectively.}

@@

Before each flight, targets used as Ground Control Points (GCPs) were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). GCPs were used to correct
georeferencing imprecision of orthomosaics with an horizontal and
vertical accuracy of 10cm. dGPS was also used to georeference quadrats
of 0.25 m² assessing the presence or absence of five key taxonomic
classes of intertidal vegetation : Bacillariophyceae (Unicellular
benthic diatoms forming biofilms at the sediment surface during low
tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf seagrass),
Chlorophyceae (green macroalgae) and Rhodophyceae (red macroalgae)
(Figure~\ref{fig-vegetation}). Pictures of each quadrat were uploaded
online to the Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}, a public open portal to store and share biodiversity
data. Each photograph was also processed to estimate the percent cover
of each type of vegetation using an image processing software (imageJ).
For all quadrats, the hyperspectral reflectance signatures of each
vegetation class was measured using an ASD FieldSpec HandHeld 2
spectroradiometer, which acquires reflectance between 325 and 1075 nm,
with 1 nm of spectral resolution. Hyperspectral signatures from quadrats
serve for dual purposes: they validate the radiometric calibration of
drone data and contribute to error reduction in photo interpretation.
The workflow developed in this study is presented in
Figure~\ref{fig-workflow}.

@@

\caption{\label{tbl-validationPX}Vegetation Classes of the model and the
number of pixels used to train and validate each class}

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model with 2 hidden layers. The training
pixels were categorized into seven different classes, representing the
various habitats encountered at the different study sites: Sediment,
Water, Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and
Rhodophyceae. Only low-altitude flights (Table~\ref{tbl-flights}) were
used for training because their 8 mm spatial resolution allowied to
avoid spectral sub-pixel mixing and to accurately identify vegetation
classes. More than 418,000 pixels at 8 mm resolution from the 3 training
flights were used to train the model (Table~\ref{tbl-validationPX}).
Twenty one variables were used by the model as predictors: the ten raw
spectral bands of the Micasense RedEdge Dual MX multispectral camera
(ranging from 444 nm to 840 nm), the same ten spectral bands
standardized using a min/max transformation (Equation~\ref{eq-std} ;
\citep{Cao2017}) and the Normalized difference vegetation index (NDVI,
Equation~\ref{eq-ndvi}). Standardisation of spectral bands is commonly
used to eliminate the scaling differences between spectra and to limit
the effect of biomass on the spectra shape \citetext{\citealp[
]{Douay2022}; \citealp{Davies2023}}.

@@

The model was applied to all flights at both 12 and 120 m of altitude.
In situ information on georeferenced class type and percent cover
collected during each flight was used to assess the model accuracy.
These images were used to construct a validation dataset indicating the
presence or absence of each class. Additionally to the quadrat-based
validation dataset, polygons of each class were photo interpreted in
order to increase the number of pixels of the validation dataset. A
confusion matrix, along with precision metrics such as global accuracy,
sensitivity, specificity, and Kappa coefficient, was generated for each
sites. All validation matrices were then aggregated to create a unique
matrix . Altogether, a total of 536,000 pixels was used to validate the
model, thus providing a geographically robust validation dataset.

@@

The high-altitude flight over Gafanha covered a total area of
approwimately 1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Spartina sp.} were
misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

Amoung the high altitude flights, the one acquired over the inner lagoon
of the Ria de Aveiro covered the largest area with approximately 1.5 km²
(Figure~\ref{fig-Boat}). This site was dominated by seagrass and the red
macroalgae \emph{Gracilaria vermicullophyla}. The classification
provided consistent results, with a patchy seagrass meadow mixed with
red macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were colonised by green
macroalgae

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The bigger the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardised reflectance spectra of each class.
Each slice represents the VIP of both raw and standardised reflectance
combined.}

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Percent cover of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal. The vertical dashed line shows the
0.85 probability of the model. Everything on the right of this line has
a probability higher then 0.85 and everything on the left of this line
has a lower probability.}

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation, as demonstrated in this study, plays a critical role in
ecological monitoring and coastal management. By distinguishing between
seagrasses and macroalgae, our approach facilitates targeted
conservation strategies, enabling more effective preservation and
restoration efforts in coastal ecosystems. The discrimination of
seagrasses from green macroalgae presents significant challenges
\citetext{\citealp[ ]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}, primarily due to
the similarities on their spectal signatures in the visible region of
the eletromagnetic spectra due to shared pigment
, Figure~\ref{fig-Pigm}, but also by the frequent spatial mixing of
these macrophytes. Besides chlorophyll-a that is common to all
vegetation types, these two green macrophytes have in common other
photosynthetic pigments such as chlorophyll-b, and accessory carotenoid
pigments such as zeaxantin, lutein and neoxanthin
(Figure~\ref{fig-Pigm}).

@@

This spectral overlap is further complicated by the intermingling of
seagrasses and green macroalgae in the same spatial locations, posing a
significant challenge for remote sensing aimed at accurately mapping and
monitoring coastal ecosystems. Our study addresses these complexities by
using high-resolution drone imagery with 10 spectral bands.
\citep{Davies2023} have shown that having at least eight spectral bands
ranging between 500 nm to 850 nm including a green band at 530 nm and
another one in the RedEdge region at 730 nm was crucial to accurately
discriminate amoung green macrophytes.

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can still be observed (Figure~\ref{fig-vegetation}).
Several factors can explain these differences such has different
concentration and/or pigment ratios \citep{bargain2013seasonal},
different cellular organisation and the disposition of the plant over
the sediment surface \citetext{\citealp[ ]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The VIP analysis of the Neural Network model (Figure~\ref{fig-VIP})
identified that the band at 531 nm was the most important one for
accurately identifying Chlorophyceae. In fact, at this wavelength,
Chlorophyceae exhibited the highest reflectance (raw and standardised)
among all other classes, highlighting the potential differences in
accessory pigment ratios (i.e.~carotenoid) between seagrasses and green
macroalgae \citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of Fucoxanthin
and Zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contributes to the lowest reflectance among all classes in the spectral
range of 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560 nm
has been identified as important for identifying this class likely due
to phycoerythrin absorption at this wavelength (\textbf{REF}). Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. This is not
a usual diagnostic wavelength for Bacillariophyceae which is mainly
identified by the presence of fucoxanthin and chlorophyll c
(Figure~\ref{fig-Pigm}), but it correspond to the absorption of
diadinoxanthin an accessory pigment characteristic of this class
\citep{meleder2003spectrometric}. Furthermore, it is the vegetation with
the lowest concentration of chlorphyll-a, pigment absorbing light both
in the blue and the red. The transparency of Bacillariophyceae makes the
reflectance of the sediment part of the overall reflectance of
Bacillariophyceae, further explaining the high reflectance in the blue.

>>>>

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
seagrass decline and habitat fragmentation. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the use of satellite data within monitoring programs. Yet,
satellite remote sensing struggles with spatial and spectral resolution,
making it difficult to discriminate seagrass from other macrophytes in
highly heterogenous meadows. Drone images at a very high spatial
resolution offer a promising solution to address challenges related to
spatial heterogeneity and intrapixel mixture. This study focuses on
using drone acquisitions with a ten spectral band sensor mirroring those
of Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12m and 120m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discriminate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. This work showed that although they
share the same pigmentary composition, seagrass and green algae can be
discriminated using a multispectral sensor with ten spectral bands at
444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. The algorithm
should now be adapted for Sentinel-2 to see if this discrimination is
still operational at a coarser spatial resolution.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between the land and oceans \citep{unsworth2022}. Seagrass meadows
provide a myriad of ecosystem services to humanity, including carbon
sequestration, oxygen production, protection against sea-level rise and
coastline erosion, and limitation of eutrophication. They serve as vital
habitats for a diverse array of marine and terrestrial species,
providing living, breeding, and feeding grounds \citetext{\citealp[
]{gardner2018}; \citealp[ ]{Zoffoli2022}; \citealp{jankowska2019}}. Due
to the concentration of human activities in coastal zones, seagrass
meadows are directly exposed to and impacted by anthropogenic pressures.
Global regression and fragmentation are currently observed due to
climate change, diseases, urbanization, land reclamation, dredging,
competition with alien species, and reduction in water quality
\citetext{\citealp[ ]{nguyen2021}; \citealp[ ]{soissons2018}; \citealp[
]{orth2006}; \citealp[ ]{lin2018}; \citealp[ ]{duffy2019}; \citealp[
]{rasheed2011long}; \citealp{chefaoui2018dramatic}}. Both habitat
fragmentation and reduction, in turn, can severely compromise the
effectiveness of ecosystem services provided by seagrass meadows. While
improvements in water quality have been recently reported in Europe,
allowing an overall recovery of seagrass ecosystems at local scale, many
coastal waters worldwide are still subjected to strong eutrophication
processes \citetext{\citealp[ ]{deSantos2019}; \citealp{Zoffoli2021}}.
Coastal eutrophication has been associated to excessive accumulation of
green macroalgae, so-called green tides (\citep{devlin2023nutrients}).
Green tides produce shade and suffocation over seagrass individuals,
thus threatening the health of seagrass ecosystems \citep{wang2022}.

@@

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a
coarse spatial resolution (\textgreater100m) over these restricted
extension areas. Missions with a high spatial resolution such as
Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral
resolution. The limited number of spectral bands is indeed a challenge
to accurately discriminate seagrass from others co-existing macrophytes.
In particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However, the application of this
approach to satellite RS remains to be validated. Moreover, patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ
measurements, due to their ability to provide spatially-explicit
observations at very high spatial resolutions (pixel size from cm to mm)
while capturing data at multi-spectral resolution \citetext{\citealp[
]{fairley2022drone}; \citealp{oh2017use}}. The versatility of drones
allows for their application across a diverse thematic range, from
coastal zone management \citetext{\citealp[ ]{adade2021}; \citealp[
]{casella2020}; \citealp{angnuureng2022}} to mapping species
distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[
]{Roman2021}; \citealp{Brunier2022Topographic}}. However, when applied
to coastal habitat mapping, most case studies are limited to a single
flight, restricting the generalizability of their application over wider
geographical scales \citetext{\citealp[ ]{Roman2021}; \citealp[
]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes, with a particular focus on discriminating
Magnoliopsida and Chlorophyceae (Seagrass and Green Algae respectively).
Ten drone flights were performed over soft-bottom intertidal areas along
two European countries (France and Portugal), covering a wide range of
habitats, from monospecific seagrass meadows to meadows mixed with
green, or red macroalgae. A deep learning algorithm was trained and
validated for macrophyte discrimination, emphasizing applicability
across diverse sites without a loss of prediction accuracy.

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourgneuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Green represents
intertidal areas.}

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take
1.2-million-pixel reflectance photographs with ten spectral bands
ranging from the blue to the near infrared (NIR): 444, 475, 531, 560,
650, 668, 705, 717, 740 and 840 nm. To ensure consistent lighting
conditions across flight paths, the drone's trajectory was aligned to
maintain a solar azimuth angle of 90 degrees. An overlap of 70\% and
80\% (side and front respectively) between each image was set for each
flight. A downwelling light sensor (DLS2) was used to acquire irradiance
data concomitantly with the camera measurements. Raw data were
calibrated in reflectance using a calibration panel reflective at
\textasciitilde50\% provided by the manufacturer. Across all sites,
flights were made at two different altitudes : 12 m or/and 120 m.
(Table~\ref{tbl-flights}).

@@

\caption{\label{tbl-flights}List of drone flight, summarizing the date,
the altitude and the purpose of each flight. 12 m and 120 m flights have
a spatial resolution of 8 and 80 mm respectively.}

@@

Before each flight, targets used as Ground Control Points (GCPs) were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). GCPs were used to correct
georeferencing imprecision of orthomosaics with an horizontal and
vertical accuracy of 10cm. dGPS was also used to georeferenced quadrats
of 0.25 m² assessing the presence or absence of five key taxonomic
classes of intertidal vegetation : Bacillariophyceae (Unicellular
benthic diatoms forming biofilms at the sediment surface during low
tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf seagrass),
Chlorophyceae (green macroalgae) and Rhodophyceae (red macroalgae)
(Figure~\ref{fig-vegetation}). Pictures of each quadrat were uploaded
online to the Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}, a public open portal to store and share biodiversity
data. Each photograph was also processed to estimate the percent cover
of each type of vegetation using an image processing software (imageJ).
For all quadrats, the hyperspectral reflectance signatures of each
vegetation class was measured using an ASD FieldSpec HandHeld 2
spectroradiometer, which acquires reflectance between 325 and 1075 nm,
with 1 nm of spectral resolution. Hyperspectral signatures from quadrats
serve for dual purposes: they validate the radiometric calibration of
drone data and contribute to error reduction in photo interpretation.
The workflow developed in this study is presented in
Figure~\ref{fig-workflow}.

@@

\caption{\label{tbl-validationPX}Vegetation Classes of the model and the
number of pixels used to train and validate each class.}

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model with 2 hidden layers. The training
pixels were categorized into seven different classes, representing the
various habitats encountered at the different study sites: Sediment,
Water, Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and
Rhodophyceae. Only low-altitude flights (Table~\ref{tbl-flights}) were
used for training because their 8 mm spatial resolution allowed to avoid
spectral sub-pixel mixing and to accurately identify vegetation classes.
More than 418,000 pixels at 8 mm resolution from the 3 training flights
were used to train the model (Table~\ref{tbl-validationPX}). Twenty one
variables were used by the model as predictors: the ten raw spectral
bands of the Micasense RedEdge Dual MX multispectral camera (ranging
from 444 nm to 840 nm), the same ten spectral bands standardized using a
min/max transformation (Equation~\ref{eq-std} ; \citep{Cao2017}) and the
Normalized difference vegetation index (NDVI, Equation~\ref{eq-ndvi}).
Standardization of spectral bands is commonly used to eliminate the
scaling differences between spectra and to limit the effect of biomass
on the spectra shape \citetext{\citealp[
]{Douay2022}; \citealp{Davies2023}}.

@@

The model was applied to all flights at both 12 and 120 m of altitude.
In situ information on georeferenced class type and percent cover
collected during each flight was used to assess the model accuracy.
These images were used to construct a validation dataset indicating the
presence or absence of each class. Additionally to the quadrat-based
validation dataset, polygons of each class were photo interpreted in
order to increase the number of pixels of the validation dataset. A
confusion matrix, along with precision metrics such as global accuracy,
sensitivity, specificity, and Kappa coefficient, was generated for each
site. All validation matrices were then aggregated to create a unique
matrix. Altogether, a total of 536,000 pixels was used to validate the
model, thus providing a geographically robust validation dataset.

@@

The high-altitude flight over Gafanha covered a total area of
approximately 1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
edges, patches of the schorre angiosperm \emph{Spartina sp.} were
misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

Among the high altitude flights, the one acquired over the inner lagoon
of the Ria de Aveiro covered the largest area with approximately 1.5 km²
(Figure~\ref{fig-Boat}). This site was dominated by seagrass and the red
macroalgae \emph{Gracilaria vermicullophyla}. The classification
provided consistent results, with a patchy seagrass meadow mixed with
red macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were colonized by green
macroalgae

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The bigger the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardized reflectance spectra of each class.
Each slice represents the VIP of both raw and standardized reflectance
combined.}

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Percent cover of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal. The vertical dashed line shows the
0.85 probability of the model. Everything on the right of this line has
a probability higher than 0.85 and everything on the left of this line
has a lower probability.}

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation, as demonstrated in this study, plays a critical role in
ecological monitoring and coastal management. By distinguishing between
seagrasses and macroalgae, our approach facilitates targeted
conservation strategies, enabling more effective preservation and
restoration efforts in coastal ecosystems. The discrimination of
seagrasses from green macroalgae presents significant challenges
\citetext{\citealp[ ]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}, primarily due to
the similarities on their spectral signatures in the visible region of
the electromagnetic spectra due to shared pigment
, Figure~\ref{fig-Pigm}, but also by the frequent spatial mixing of
these macrophytes. Besides chlorophyll-a that is common to all
vegetation types, these two green macrophytes have in common other
photosynthetic pigments such as chlorophyll-b, and accessory carotenoid
pigments such as zeaxantin, lutein and neoxanthin
(Figure~\ref{fig-Pigm}).

@@

This spectral overlap is further complicated by the intermingling of
seagrasses and green macroalgae in the same spatial locations, posing a
significant challenge for remote sensing aimed at accurately mapping and
monitoring coastal ecosystems. Our study addresses these complexities by
using high-resolution drone imagery with 10 spectral bands.
\citep{Davies2023} have shown that having at least eight spectral bands
ranging between 500 nm to 850 nm including a green band at 530 nm and
another one in the RedEdge region at 730 nm was crucial to accurately
discriminate among green macrophytes.

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can still be observed (Figure~\ref{fig-vegetation}).
Several factors can explain these differences such has different
concentration and/or pigment ratios \citep{bargain2013seasonal},
different cellular organization and the disposition of the plant over
the sediment surface \citetext{\citealp[ ]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The VIP analysis of the Neural Network model (Figure~\ref{fig-VIP})
identified that the band at 531 nm was the most important one for
accurately identifying Chlorophyceae. In fact, at this wavelength,
Chlorophyceae exhibited the highest reflectance (raw and standardized)
among all other classes, highlighting the potential differences in
accessory pigment ratios (i.e.~carotenoid) between seagrasses and green
macroalgae \citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of Fucoxanthin
and Zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contributes to the lowest reflectance among all classes in the spectral
range of 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560 nm
has been identified as important for identifying this class likely due
to phycoerythrin absorption at this wavelength (\textbf{REF}). Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. This is not
a usual diagnostic wavelength for Bacillariophyceae which is mainly
identified by the presence of fucoxanthin and chlorophyll c
(Figure~\ref{fig-Pigm}), but it correspond to the absorption of
diadinoxanthin an accessory pigment characteristic of this class
\citep{meleder2003spectrometric}. Furthermore, it is the vegetation with
the lowest concentration of chlorophyll-a, pigment absorbing light both
in the blue and the red. The transparency of Bacillariophyceae makes the
reflectance of the sediment part of the overall reflectance of
Bacillariophyceae, further explaining the high reflectance in the blue.
